A Simplified Spectral Deferred Correction Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2020-08-11 simplified_sdc_intro/paper.tex
new title
some more fixet s

\begin{document}
%======================================================================
% Title
%======================================================================
%\title{A Simple Spectral Deferred Correction Method to Couple Nuclear Burning and Hydrodynamics for Detonations}

\title{So you've decided to detonate helium}

\shorttitle{Detonations and NSE}
\shortauthors{}

@@

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Helium detonations can be challenging to model because of the
difficulty in accurately coupling hydrodynamics and reactions.  This
is particularly acute in the ash of the detonation where nuclear
statistical equilibrium is established.  We look a several different
methods of coupling explicit hydrodynamics and stiff reaction sources,
applied to detonations which reach NSE conditions.  This is all done
in the framework of the \castro\ hydrodynamics code, and all algorithm
implementations are freely available.
\end{abstract}

@@

A particular difficult phase of evolution to model is the nuclear
statistic equilibrium that sets in with helium burning for
temperatures in excess of $4\times 10^9~\mathrm{K}$.  Physically, the
forward and reverse rates should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  This large change in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep, but this makes simulations prohibitively expensive.
Sometimes the burning is simply halted on a zone-by-zone basis when
NSE conditions are reached (e.g., as in \citealt{hedet}).
Alternately, at high temperatures, a reaction network can be replaced
with a table of NSE abundances and the zone's composition set through
table look-ups\MarginPar{example? GCD?}.

In this paper, we look at some different approaches to coupling
hydrodynamics and reactions, using a helium detonation that attains
NSE as our testbed.  We use the \castro\ hydrodynamics code for all of
our numerical experiments, and all of the code to reproduce the results
in this paper are in the \castro\ github repository.

@@

We solve the Euler equations for compressible, reacting flow.  We'll focus on
one-dimensional flow, and the system appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where $\Uc = (\rho, (\rho X_k), (\rho u), (\rho E))^\intercal$ are
the conserved fluid quantities: mass density, $\rho$, velocity, $u$,
specific total energy, $E$, and nuclear
species mass fractions, $X_k$.  The specific total energy relates to
the specific internal energy as $E = e + u^2/2$.  The
corresponding fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\ \rho X_k u \\ \rho u u + p \\  \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state,
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

Finally, $\Sc(\Uc)$ are the source terms, which we decompose into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc)
\end{equation}
In this paper, we will not consider any hydrodynamical sources, but we
will carry the term $\Hb$ around for completeness.  The reactive
sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\ \rho \omegadot_k \\ 0 \\ \rho \dot{S}
  \end{array} \right )
\end{equation}
where $\omegadot_k$ is the creation rate for species $k$ and $\dot{S}$
is the energy generation rate per unit mass.

@@

We use the \castro\ hydrodynamics code~\cite{castro}, together with
the corner transport upwind piecewise parabolic method
hydrodynamics~\citep{millercolella:2002}.  This is a finite-volume
method that uses characteristic tracing to predict a time-centered
flux through the interfaces of the grid zones.  For pure
hydrodynamics, these time-centered fluxes result in second-order
accurate time integration.  With reactions, we want to couple
hydrodynamics and the reaction sources to second order.  Nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting is traditionally employed
here, and our goal in this paper is to look at several different
approaches for splitting.

We describe several time-integration techniques for coupling
hydrodynamics and reactions in the next sections.

@@

@@

@@

\subsection{Balanced Splitting}

@@

      Note that our advection terms are piecewise constant \MarginPar{need to think about this more}
      approximations, so when we integrate our system, their
      contributions will be linear in time.  Since $(\rho X_k)$ sees a
      linear advective term, $\rho$ does as well.  $(\rho \Ub)$ and
      $(\rho E)$ will also have linear contributions in time from
      advection.  We can see that the internal energy will also have a
      linear contribution by recognizing that the kinetic energy term,
      $K = (\rho |\Ub|^2)/\rho$ is linear in time.
      Since the reactions don't affect $\rho$ and $(\rho \Ub)$, we can
      algebraically update these using this piecewise linear behavior:
      \begin{align}
        \rho(t) &= \rho^n + \Adv{\rho}^{n+1/2,(k)} \, t \\
        (\rho \Ub)(t) &= (\rho \Ub)^n + \Adv{(\rho \Ub)}^{n+1/2,(k)} \, t
      \end{align}
      At the end of the integration, our new state is $\Uc^{n+1,(k)}$.

@@

\subsection{Old Primitive Stuff}

and our primitive system as:
\begin{equation}
\qb_t = \Advs{\qb} + \Rb(\qb)
\end{equation}
with
\begin{equation}
\Advs{\qb} = -\Ab^\xv(\qb) \qb_x - \Ab^\yv(\qb) \qb_y - \Ab^\zv(\qb) \qb_z + \Sqhydro
\end{equation}

>>>>

\begin{document}
%======================================================================
% Title
%======================================================================
\title{A Simplified Spectral Deferred Correction Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium}

\shorttitle{A Simplified SDC Method}
\shortauthors{}

@@

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute at high temperatures where nuclear
statistical equilibrium is established.  We develop a simplified
approach to spectral deferred corrections coupling of explicit
hydrodynamics and stiff reaction sources and apply it to some example
problems.  This is all done in the framework of the
\castro\ hydrodynamics code, and all algorithm implementations are
freely available.
\end{abstract}

@@

A particular difficult phase of evolution to model is the nuclear
statistic equilibrium that sets in with helium burning for
temperatures in excess of $4\times 10^9~\mathrm{K}$.  Physically, the
forward and reverse rates should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  This large change in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep, but this makes simulations prohibitively expensive.
Sometimes the burning is simply halted on a zone-by-zone basis when
NSE conditions are reached (e.g., as in \citealt{hedet}).
Alternately, at high temperatures, a reaction network can be replaced
with a table of NSE abundances and the zone's composition set through
table look-ups (e.g.\ \citealt{ma:2013}) \MarginPar{need more here}

In \cite{castro_sdc}, we developed second- and fourth-order accurate
methods in space and time for coupling hydrodynamics and nuclear
reaction networks and demonstrated the method of a variety of test
problems.  The approach her is considerably simpler, reusing the main
CTU hydrodynamics construction and a largely similar ODE integration
scheme, making this method easier to add to existing simulation codes.
However, it is restricted to second-order accuracy overall.  We term this
algorithm the ``simplified SDC method''.
We use the \castro\ hydrodynamics code for all of
our numerical experiments, and all of the code to reproduce the results
in this paper are in the \castro\ github repository.

@@

We solve the Euler equations for compressible, reacting flow.  We'll focus on
one-dimensional flow for this description, and the system appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity, $u$,
specific total energy, $E$, and nuclear
species mass fractions, $X_k$.  The specific total energy relates to
the specific internal energy as $E = e + u^2/2$.  The
corresponding fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\ \rho X_k u \\ \rho u u + p \\  \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state,
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

Finally, $\Sc(\Uc)$ are the source terms, which we decompose into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc)
\end{equation}
 The reactive
sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\ \rho \omegadot_k \\ 0 \\ \rho \dot{S}
  \end{array} \right )
\end{equation}
where $\omegadot_k$ is the creation rate for species $k$ and $\dot{S}$
is the energy generation rate per unit mass.

@@

We use the \castro\ hydrodynamics code~\cite{castro}, together with
the corner transport upwind piecewise parabolic method
hydrodynamics~\citep{millercolella:2002}.  This is a finite-volume
method that uses characteristic tracing to predict a time-centered
flux through the interfaces of the grid zones.  For pure
hydrodynamics, these time-centered fluxes result in second-order
accurate time integration.  With reactions, we want to couple
hydrodynamics and the reaction sources to second order.  Nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting is traditionally employed
here, and we compare to it in this paper.

@@

In Strang splitting, we first evolve the system without advection
through $\Delta t/2$, then do the advective update without reactions
through $\Delta t$, and finally do the last reactive update through $\Delta t/2$.

In the absence of advective terms, our system appears as:
\begin{align}
\odt{\rho} & = 0 \\
\odt{(\rho X_k)} &= \rho \omegadot_k \\
\odt{(\rho u)} &= 0 \\
\odt{(\rho E)} &= \rho \Sdot 
\end{align}
Ww can write the energy equation as:
\begin{equation}
\odt{(\rho E)} = \odt{(\rho e)} + \odt{K} = \rho \Sdot
\end{equation}
where $K$ is the kinetic energy, $K = |u|^2/2$.  Since the velocity is
unchanged by reactions, this means that
\begin{equation}
\odt{(\rho e)} = \rho \odt{e} = \rho \Sdot
\end{equation}
The reaction rates are typically expressed as $\omegadot_k(\rho, T,
X_k)$, if we evolve this system.  This requires an EOS inversion each
time we need to evaluate the reactive terms.

Normally we want to integrate temperature temperature.  We can write the 
energy equation as:
\begin{equation}
\rho \odt{e} = \rho \left [ \underbrace{\left . \frac{\partial e}{\partial T} \right |_\rho}_{c_v}
                            \odt{T} +
                            \left . \frac{\partial e}{\partial \rho} \right |_T
                            \cancelto{0}{\odt{\rho}} \;\; \right ] = \rho \Sdot
\end{equation}
reducing to
\begin{equation}
\odt{T} = \frac{1}{c_v} \Sdot
\end{equation}
We can alternately derive this from an enthalpy equation, which is the correct
formulation when we are at constant pressure, and we get the same temperature
equation but with $c_v$ replaced by $c_p$.
Technically this requires an EOS call to get $c_x$, but if the
specific heat varies slowly, we can freeze it during the reaction
solve.

@@

write these out

@@

@@

      Note that our advection terms are piecewise constant \MarginPar{need to think about this more}
      approximations, so when we integrate our system, their
      contributions will be linear in time.  Since $(\rho X_k)$ sees a
      linear advective term, $\rho$ does as well.  $(\rho \Ub)$ and
      $(\rho E)$ will also have linear contributions in time from
      advection.  We can see that the internal energy will also have a
      linear contribution by recognizing that the kinetic energy term,
      $K = (\rho |\Ub|^2)/\rho$ is linear in time.
      Since the reactions don't affect $\rho$ and $(\rho \Ub)$, we can
      algebraically update these using this piecewise linear behavior:
      \begin{align}
        \rho(t) &= \rho^n + \Adv{\rho}^{n+1/2,(k)} \, t \\
        (\rho \Ub)(t) &= (\rho \Ub)^n + \Adv{\rho \Ub}^{n+1/2,(k)} \, t
      \end{align}
      At the end of the integration, our new state is $\Uc^{n+1,(k)}$.

@@

\subsection{Nuclear Statistical Equilibrium}
